---
title: "Whole genome sequencing of methicillin-resistant <i>Staphylococcus aureus<i> in wastewater and clinical samples
"
author: "Julia Kamer and Sheena Conforti"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
    code_folding: hide  # or "show" if you want it open by default
  bookdown::html_document2:
    toc: true
    toc_depth: 2
    code_folding: hide  # hides code by default, adds toggle
    number_sections: false
    df_print: paged
  pdf_document:
    keep_tex: true
---

```{r setup, include=FALSE}
# suppress warnings and messages to create neat final output
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

------------------------------------------------------------------------

# 1. Overview

This R Markdown notebook documents the full analysis pipeline for the genomic diversity, antibiotic resistance gene content, virulence factor content, and phylogenetic relatedness of 115 methicillin-resistant *Staphylococcus aureus* (MRSA) isolates (89 clinical, 26 wastewater) from Switzerland, supporting the main manuscript and its supplementary material. Table/figure numbers below refer to the current final versions of the main text and supplementary material. The workflow includes:

-   Sequencing quality and isolate inclusion (Section 3): reconciles the per-sample sequencing/QC log into the final analysis set, reported in Table S5.

-   Genomic diversity (Section 4): sequence type (ST), SCCmec type, and spa type, and their combinations, characterized by source (Table S1, Table S6, Table S7, Table S8, Table S9). These counts, together with the cgSNP relatedness counts from Section 7, are compiled by hand into Table 1 of the main manuscript (see note below).

-   Antibiotic resistance genes, ARGs (Section 5): per-gene and per-class carriage, rarefaction-adjusted richness comparisons between sources, and multidrug resistance (Table S10, Table S11, Figure 2, Figure S2, Figure S3).

-   Virulence factor genes, VFs (Section 6): per-gene carriage including Panton-Valentine leukocidin (PVL) status (Table S12), and isolate-level category summaries. The resistance-class and median-ARG-count summaries from Section 5 and the VF category summary from Section 6 are compiled by hand into Table 2 of the main manuscript (see note below).

-   Genetic relatedness (Section 7): core-genome SNP (cgSNP)-based collapsing of identical isolates from the same sample (Table S2), and identification of closely related (≤15 cgSNP) isolate pairs within and between source compartments (isolate-pair counts feed Table 1; the pair-level listing is exported separately as TableSX).

-   Geographic context (Section 8): map of the six wastewater treatment plant (WWTP) catchment areas and the hospitals that contributed clinical isolates (Figure S1).

**Not generated by this code.** Table S3 (list of antimicrobial resistance genes and their associated antibiotic classes, from CARD) and Table S4 (number of presumptive MRSA isolates collected per WWTP by sampling date) are additional datasets assembled outside this pipeline; no chunk in this notebook produces them.

**Table 1 and Table 2 are compiled, not generated, by a single chunk.** Table 1 (isolate/ST/SCCmec/spa counts plus cgSNP relatedness) draws on several non-contiguous chunks in Sections 4 and 7; Table 2 (resistance classes, ARG/class medians, and virulence factor categories) draws on non-contiguous chunks in Sections 5 and 6. Each contributing chunk is flagged in-line (e.g. "part of Table 1/2") where it occurs, and its output is a partial table, not the final published one.

**Figure 1 (the phylogenetic tree) is not rendered by this notebook.** The tree itself is built externally in iTOL from the core-genome alignment and finished in Inkscape. This notebook does, however, generate the ST, SCCmec, spa, source, and PVL-positivity data used to annotate that tree, including the ≥80% identity, both-genes rule (Section 6.2) that determines which isolates receive the red PVL marker in Figure 1.

------------------------------------------------------------------------

# 2. Load packages

Loads every package used downstream: tidyverse/dplyr/tidyr for data wrangling, kableExtra for table rendering, sf for the spatial data in Section 8, igraph for the isolate-pair network in Section 7, rstatix/FSA/ggsignif for the Mann-Whitney tests in Section 5, and ape/phangorn (retained for compatibility with earlier tree-adjacent steps, though the tree itself is built externally — see Section 1 Overview). Loaded up front so a missing dependency surfaces immediately rather than mid-pipeline.

If not yet installed, install them using \`install.packages("package_name")\`.

```{r load-packages, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(gridExtra)
library(ggpubr)
library(tidyr)
library(seqinr)
library(readr)
library(ggthemes)
library(patchwork)
library(kableExtra)
library(FSA)
library(tibble)
library(tidyverse)
library(RColorBrewer)
library(reshape2)
library(igraph)
library(phangorn)
library(ape)
library(readxl)
library(pheatmap)
library(rstatix)
library(vegan)
library(ggsignif)
library(sf)
library(stringr)
library(ggspatial) 
library(scales) 
```

------------------------------------------------------------------------

# 3. Sequencing results

Reconciles the raw per-sample sequencing/QC log with the final isolate set, producing the accounting (Table S5) for why 115 of the 117 candidate isolates were retained in the final analysis.

## 3.1 Load data

Reads the per-sample QC/sequencing log (batch, contamination flag, coverage, assembly quality, final-tree inclusion) that every count in this section is derived from.

```{r load-sequencing, warning=FALSE, error=FALSE, eval=TRUE}
sequencing <- read.csv("data/sequencing.csv", header =TRUE)
```

## 3.2 Sequencing output summary (tabS5)

Collapses the per-sample log into the isolate-level counts reported in Table S5 (totals by source, tree inclusion, exclusion reasons, negative controls) — the QC accounting behind the final n = 115 analysis set used everywhere else in this notebook.

```{r tableS5, warning=FALSE, message=FALSE, eval=TRUE}
# STEP 0: Standardize naming
sequencing <- sequencing %>%
  mutate(
    Batch = str_trim(Batch),
    source = case_when(source == "clinical" ~ "clinics", TRUE ~ source)
  )

# STEP 1: Filter and deduplicate only ESBL-E. coli samples
mrsa <- sequencing %>%
  filter(sample_type == "mrsa") %>%
  select(Sample, Batch, source, Contamination, low_coverage, low_quality_assembly, Final.tree) %>%
  distinct()

# STEP 2: Per-sample summary
mrsa_summary <- mrsa %>%
  group_by(Sample) %>%
  summarise(
    Contamination_flag = any(Contamination == "yes"),
    low_coverage_flag = any(low_coverage == "yes"),
    low_quality_flag = any(low_quality_assembly == "yes"),
    any_clean = any(Contamination == "no" & low_coverage == "no" & low_quality_assembly == "no"),
    in_final_tree = any(Final.tree == "yes"),
    source = first(source),
    .groups = "drop"
  ) %>%
  mutate(
    excluded = !any_clean,
    exclusion_reason = case_when(
      excluded & Contamination_flag ~ "Due to contamination",
      excluded & low_coverage_flag ~ "Due to low coverage",
      excluded & low_quality_flag ~ "Due to low assembly quality",
      TRUE ~ NA_character_
    )
  )

# STEP 3: Summary counts
total_n <- nrow(mrsa_summary)

total_by_source <- mrsa_summary %>%
  dplyr::count(source) %>%
  rename(total = n)

final_tree_by_source <- mrsa_summary %>%
  filter(in_final_tree) %>%
  dplyr::count(source) %>%
  rename(in_final_tree = n)

excluded_total <- mrsa_summary %>%
  filter(excluded) %>%
  nrow()

excluded_by_reason <- mrsa_summary %>%
  filter(excluded) %>%
  dplyr::count(exclusion_reason) %>%
  rename(reason = exclusion_reason, n = n)

# STEP 4: QC samples by type and batch (correct strain-level)
neg_controls <- sequencing %>%
  filter(sample_type == "negative_control") %>%
  dplyr::count(Batch) %>%
  rename(n_neg_controls = n)

# Count specific strain types
count_strain_batch <- function(strain_name) {
  sequencing %>%
    filter(sample_type == strain_name) %>%
    dplyr::count(Batch) %>%
    complete(Batch = c("batch1", "batch2"), fill = list(n = 0)) %>%
    pull(n)
}

# Individual strain counts by batch
ecoli1_b1 <- count_strain_batch("ecoli_control_strain1")[1]
ecoli2_b1 <- count_strain_batch("ecoli_control_strain2")[1]


# STEP 5: Resequencing outcomes
low_cov_batch1 <- sequencing %>%
  filter(sample_type == "esblecoli", low_coverage == "yes", Batch == "batch1") %>%
  pull(Sample) %>%
  unique()

rescued_in_batch2 <- sequencing %>%
  filter(sample_type == "esblecoli", Batch == "batch2", Sample %in% low_cov_batch1,
         low_coverage == "no", Contamination == "no", low_quality_assembly == "no") %>%
  pull(Sample) %>%
  unique()

failed_both_batches <- setdiff(low_cov_batch1, rescued_in_batch2)

# Safe accessor helper
get_count <- function(df, var, value) {
  result <- df %>% filter({{ var }} == value) %>% pull(2)
  if (length(result) == 0) return(0) else return(result)
}

# STEP 6: Build final summary table
tableS5 <- tibble(
  Category = c(
    "Total unique MRSA isolates",
    "From wastewater",
    "From clinics",
    "Included in final tree (total)",
    "Included in final tree (wastewater)",
    "Included in final tree (clinics)",
    "Excluded from final tree (total)",
    "Exclusion due to contamination",
    "Exclusion due to low coverage",
    "Exclusion due to low assembly quality",
    "AE negative control samples",
    "E. coli strain1",
    "E. coli strain2"
  ),
  N = c(
    total_n,
    get_count(total_by_source, source, "wastewater"),
    get_count(total_by_source, source, "clinics"),
    sum(mrsa_summary$in_final_tree),
    get_count(final_tree_by_source, source, "wastewater"),
    get_count(final_tree_by_source, source, "clinics"),
    excluded_total,
    get_count(excluded_by_reason, reason, "Due to contamination"),
    get_count(excluded_by_reason, reason, "Due to low coverage"),
    get_count(excluded_by_reason, reason, "Due to low assembly quality"),
    get_count(neg_controls, Batch, "batch1"),
    ecoli1_b1, ecoli2_b1
  )
)

tableS5$N[is.na(tableS5$N)] <- 0

# STEP 7: Display
tableS5 %>%
  kable(
    format = "html",
    longtable = TRUE,
    booktabs = TRUE,
    caption = "Table S5: Comprehensive summary of MRSA sequencing outcomes, quality filtering, and control types by strain."
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "repeat_header"),
    full_width = FALSE,
    position = "center"
  )

# Save the final sequencing summary table
#write.csv(tableS5, "./tables/TableS5.csv", row.names = FALSE)
```

------------------------------------------------------------------------

# 4. MRSA diversity

Characterizes the genomic diversity of the final isolate set — sequence type, SCCmec type, spa type, and their combinations — stratified by source. These tables (Table S1, S6-S9) supply the isolate/ST/SCCmec/spa counts that are compiled by hand into Table 1 of the main manuscript (see Section 1 Overview).

## 4.1 Load metadata

Loads the per-isolate metadata table (source, location, dates, ST/SCCmec/spa calls) that every table in this section is built from.

```{r upload-metadata, warning=FALSE, message=FALSE, eval=TRUE}
sample_metadata <- read.csv("data/sample_annotations.csv", header=TRUE)
```

## 4.2 Sample summary (tabS1)

Produces Table S1: sampling location and collection-period range by source, establishing where and when isolates were collected.

```{r tableS1, eval=TRUE}
# Table S1: Sampling locations and collection period
tableS1 <- sample_metadata %>%
  group_by(Source, Location) %>%
  summarise(
    `No. isolates` = n(),
    `First sample` = min(as.Date(Date, format = "%d_%m_%Y"), na.rm = TRUE),
    `Last sample` = max(as.Date(Date, format = "%d_%m_%Y"), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(Source, Location)

# Display Table S1
tableS1 %>%
  kable(
    format = "html",
    longtable = TRUE,
    booktabs = TRUE,
    escape = FALSE,
    caption = paste0(
      "Table S1: Sampling locations and collection period of methicillin-resistant ",
      "<i>Staphylococcus aureus</i> (MRSA) isolates included in this study. ",
      "The table reports the number of isolates recovered from each source and location, ",
      "together with the first and last sampling dates."
    )
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "repeat_header"),
    full_width = FALSE,
    position = "center"
  )

# Save to file
#write.csv(tableS1, "./tables/TableS1.csv", row.names = FALSE)
```

## 4.3 Sequence type (ST) diversity (tabS6)

Produces Table S6: the ST distribution overall and by source, the primary measure of clonal diversity referenced in the Results.

```{r tableS6, message=FALSE, warning=FALSE, eval=TRUE}
# Count total isolates per source for denominator
source_totals <- sample_metadata %>%
  filter(ST != "", ST != "failed") %>%
  dplyr::count(Source, name = "total_n") %>%
  pivot_wider(names_from = Source, values_from = total_n)

total_n <- nrow(sample_metadata %>% filter(ST != "", ST != "failed"))

# Generate Table S6: Sequence type distribution by source
tableS6 <- sample_metadata %>%
  filter(!is.na(ST), ST != "failed") %>%
  group_by(ST) %>%
  dplyr::summarise(
    total = n(),
    clinical = sum(Source == "clinics"),
    wastewater = sum(Source == "wastewater"),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    percent_total = round(100 * total / total_n, 1),
    percent_clinical = round(100 * clinical / source_totals$clinics, 1),
    percent_wastewater = round(100 * wastewater / source_totals$wastewater, 1),
    `Number (percentage %) of total isolates` = paste0(total, " (", percent_total, "%)"),
    `Number (percentage %) of clinical isolates` = paste0(clinical, " (", percent_clinical, "%)"),
    `Number (percentage %) of wastewater isolates` = paste0(wastewater, " (", percent_wastewater, "%)")
  ) %>%
  arrange(desc(total)) %>%
  select(
    SequenceType = ST,
    `Number (percentage %) of total isolates`,
    `Number (percentage %) of clinical isolates`,
    `Number (percentage %) of wastewater isolates`,
  )

# Display Table S6
tableS6 %>%
  kable(
    format = "html",
    longtable = TRUE,
    booktabs = TRUE,
    escape = FALSE,
    caption = "Table S6: Distribution of methicillin-resistant <i>Staphylococcus aureus</i>  (MRSA) isolates by sequence type (ST), identified by the analysis of seven housekeeping genes of the multilocus sequence typing (MLST) Achtman scheme using MLST v.2.16.1 (https://github.com/tseemann/mlst). The table includes the number and percentage of isolates for each ST. Sequence types not determined are assigned na."
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "repeat_header"),
    full_width = FALSE,
    position = "center"
  )

# Save to file
#write.csv(tableS6, "./tables/TableS6.csv", row.names = FALSE)
```

## 4.4 SCCmec diversity (tabS7)

Produces Table S7: SCCmec type distribution by source, used together with ST and spa type to characterize strain background. (Renumbered from the original duplicate "4.2" heading to keep subsection numbers sequential.)

```{r tableS7, warning=FALSE, message=FALSE, eval=TRUE}
# Table S7 generation
tableS7 <- sample_metadata %>%
  filter(SCCmec != "", SCCmec != "failed") %>%
  group_by(SCCmec) %>%
  summarise(
    total = n(),
    clinical = sum(Source == "clinics"),
    wastewater = sum(Source == "wastewater"),
    .groups = "drop"
  ) %>%
  mutate(
    percent_total = round(100 * total / total_n, 1),
    percent_clinical = round(100 * clinical / source_totals$clinics, 1),
    percent_wastewater = round(100 * wastewater / source_totals$wastewater, 1),
    `Number (percentage %) of total isolates` = paste0(total, " (", percent_total, "%)"),
    `Number (percentage %) of clinical isolates` = paste0(clinical, " (", percent_clinical, "%)"),
    `Number (percentage %) of wastewater isolates` = paste0(wastewater, " (", percent_wastewater, "%)")
  ) %>%
  select(
    SCCmec,
    `Number (percentage %) of total isolates`,
    `Number (percentage %) of clinical isolates`,
    `Number (percentage %) of wastewater isolates`,
  ) %>%
  arrange(SCCmec)

# Display Table S7
tableS7 %>%
  kable(
    format = "html",
    longtable = TRUE,
    booktabs = TRUE,
    escape = FALSE,
    caption = "Table S7: Distribution of methicillin-resistant <i>Staphylococcus aureus</i>  (MRSA) isolates by SCCmec type, stratified by source. SCCmec types not determined are assigned na."
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "repeat_header"),
    full_width = FALSE,
    position = "center"
  )

# Save to file
#write.csv(tableS7, "./tables/TableS7.csv", row.names = FALSE)

```

## 4.5 spa diversity (tabS8)

Produces Table S8: spa type distribution by source, the finer-resolution typing marker used to identify shared clones between clinical and wastewater isolates. (Renumbered from the original duplicate "4.3" heading.)

```{r tableS8, warning=FALSE, message=FALSE, eval=TRUE}
# Table S8 generation
tableS8 <- sample_metadata %>%
  filter(spa != "", spa != "failed") %>%
  group_by(spa) %>%
  summarise(
    total = n(),
    clinical = sum(Source == "clinics"),
    wastewater = sum(Source == "wastewater"),
    .groups = "drop"
  ) %>%
  mutate(
    percent_total = round(100 * total / total_n, 1),
    percent_clinical = round(100 * clinical / source_totals$clinics, 1),
    percent_wastewater = round(100 * wastewater / source_totals$wastewater, 1),
    `Number (percentage %) of total isolates` = paste0(total, " (", percent_total, "%)"),
    `Number (percentage %) of clinical isolates` = paste0(clinical, " (", percent_clinical, "%)"),
    `Number (percentage %) of wastewater isolates` = paste0(wastewater, " (", percent_wastewater, "%)")
  ) %>%
  select(
    spa,
    `Number (percentage %) of total isolates`,
    `Number (percentage %) of clinical isolates`,
    `Number (percentage %) of wastewater isolates`,
  ) %>%
  arrange(spa)

# Display Table S8
tableS8 %>%
  kable(
    format = "html",
    longtable = TRUE,
    booktabs = TRUE,
    escape = FALSE,
    caption = "Table S8: Distribution of methicillin-resistant Staphylococcus aureus (MRSA) isolates by spa type, stratified by source. Spa types not determined by either method (WGS or Sanger) are assigned na."
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "repeat_header"),
    full_width = FALSE,
    position = "center"
  )

# Save to file
#write.csv(tableS8, "./tables/TableS8.csv", row.names = FALSE)
```

## 4.6 ST-SCCmec-spa combinations (tabS9)

Produces Table S9: the count of unique ST-SCCmec-spa combinations, summarizing how much of the observed diversity reflects distinct genetic backgrounds versus repeated clones. (Renumbered to follow 4.5 sequentially.)

```{r tableS9, eval=TRUE}
# Generate the table of all fully assigned ST-SCCmec-spa combinations
tableS9 <- sample_metadata %>%
  # Convert variables to character and remove surrounding spaces
  mutate(
    ST = trimws(as.character(ST)),
    SCCmec = trimws(as.character(SCCmec)),
    spa = trimws(as.character(spa))
  ) %>%
  
  # Exclude isolates where at least one assignment is missing or recorded as "NA"
  filter(
    !is.na(ST),
    !is.na(SCCmec),
    !is.na(spa),
    ST != "",
    SCCmec != "",
    spa != "",
    !tolower(ST) %in% c("na", "n/a"),
    !tolower(SCCmec) %in% c("na", "n/a"),
    !tolower(spa) %in% c("na", "n/a")
  ) %>%
  
  # Create the combined genotype
  mutate(
    combination = paste(ST, SCCmec, spa, sep = "-")
  ) %>%
  
  # Count isolates by combination and source
  dplyr::count(combination, Source, name = "n") %>%
  
  # Create separate columns for clinical and wastewater isolates
  pivot_wider(
    names_from = Source,
    values_from = n,
    values_fill = 0
  ) %>%
  
  # Calculate totals and percentages among fully assigned isolates
  mutate(
    total = clinics + wastewater,
    total_percent = round(100 * total / sum(total), 1),
    clinical_percent = round(100 * clinics / sum(clinics), 1),
    wastewater_percent = round(
      100 * wastewater / sum(wastewater),
      1
    ),
    total_display = paste0(total, " (", total_percent, "%)"),
    clinical_display = paste0(
      clinics,
      " (",
      clinical_percent,
      "%)"
    ),
    wastewater_display = paste0(
      wastewater,
      " (",
      wastewater_percent,
      "%)"
    )
  ) %>%
  
  # Sort and retain final display columns
  arrange(desc(total), combination) %>%
  select(
    `ST-SCCmec-spa combination` = combination,
    `Total isolates, n (%)` = total_display,
    `Clinical isolates, n (%)` = clinical_display,
    `Wastewater isolates, n (%)` = wastewater_display
  )

# Count the number of distinct fully assigned combinations
n_combinations <- nrow(tableS9)

# Display the table
tableS9 %>%
  kable(
    format = "html",
    booktabs = TRUE,
    align = c("l", "c", "c", "c"),
    caption = paste0(
      "Table S9: Distribution of ",
      n_combinations,
      " unique ST-SCCmec-spa combinations among MRSA isolates. ",
      "Isolates with undefined ST, SCCmec, or spa type assignments were excluded."
    )
  ) %>%
  kable_styling(
    full_width = FALSE,
    bootstrap_options = c("striped", "condensed")
  )

#Save the table
#write.csv(tableS9,"./tables/TableS9.csv", row.names = FALSE)
```

------------------------------------------------------------------------

# 5. Antibiotic Resistant Genes (ARGs)

Characterizes antimicrobial resistance gene content: which ARGs are present, how many/which resistance classes they confer, multidrug resistance status, and whether wastewater and clinical isolates differ. Feeds Table 2, Figure 2, Figure S2, Figure S3, Table S10, and Table S11.

## 5.1 Load ARGs data

Loads the raw Abricate ARG hits per isolate and the gene-to-antibiotic-class lookup table used to translate individual genes into resistance classes.

```{r import-args-data, message=FALSE, warning=FALSE, eval=TRUE}
arg_data <- read.delim("data/args_abricate.txt", header = TRUE, check.names = FALSE)
arg_class <- read.csv("data/gene_and_class.csv", header =TRUE)
```

## 5.2 Format ARGs data

Cleans sample identifiers and joins in Source/Sample_date so ARG presence can be linked back to the same per-isolate metadata used in Section 4.

```{r format-args-data, message=FALSE, warning=FALSE, eval=TRUE}

# Rename the first column to "Sample"
colnames(arg_data)[1] <- "Sample"

# Extract the clean sample names (between last "/" and first "_")
arg_data$Sample <- gsub("^.*/|_.*$", "", arg_data$Sample)

# Replace "." with NA in gene columns to handle missing values
arg_data[arg_data == "."] <- NA

#Replace sampling names 
arg_data <- arg_data %>%
  rename(Sample_ID = Sample) %>%  # Step 1: Rename technical ID
  left_join(
    sample_metadata %>% select(Sample_ID, Source, Sample_date),
    by = "Sample_ID"
  ) %>%
  rename(Sample = Sample_date) %>%  # Step 2: Use human-readable name
  relocate(Sample, Source) %>%      # Optional: put Sample and Source first
  select(-Sample_ID)                # Optional: drop technical ID
```

## 5.3 Antibiotic Resistant Genes

Groups the four analyses that describe ARG burden and its relationship to source: per-gene frequencies (Table S11), rarefaction-adjusted richness comparisons (Figure S3), and the resistance-class summary that is one of three chunks compiled into Table 2. (Renumbered from "5.4" to close the gap left by the missing "5.3" in the original numbering.)

### 5.3.1 Count genes in isolates by source (tabS11)

Produces Table S11: the carriage frequency of every individual ARG, overall and by source — the most granular resistance-gene inventory reported.

```{r tableS11, message=FALSE, warning=FALSE, eval=TRUE}
# ARG frequency summaries overall and by source
# Total number of isolates
total_isolates <- nrow(arg_data)

# ARG columns only
arg_gene_cols <- setdiff(
  colnames(arg_data),
  c("Sample", "Source", "NUM_FOUND")
)

# Number of unique ARGs detected overall and by source
n_unique_args_overall <- arg_data %>%
  select(all_of(arg_gene_cols)) %>%
  summarise(across(everything(), ~ any(!is.na(.)))) %>%
  unlist() %>%
  sum()

n_unique_args_clinical <- arg_data %>%
  filter(Source == "clinics") %>%
  select(all_of(arg_gene_cols)) %>%
  summarise(across(everything(), ~ any(!is.na(.)))) %>%
  unlist() %>%
  sum()

n_unique_args_wastewater <- arg_data %>%
  filter(Source == "wastewater") %>%
  select(all_of(arg_gene_cols)) %>%
  summarise(across(everything(), ~ any(!is.na(.)))) %>%
  unlist() %>%
  sum()

# 1) Overall ARG summary for downstream analyses
gene_counts <- arg_data %>%
  summarise(
    across(
      all_of(arg_gene_cols),
      ~ sum(!is.na(.))
    )
  )

gene_percentages <- gene_counts / total_isolates * 100

gene_arg_summary <- tibble(
  Gene = colnames(gene_counts),
  Count = as.numeric(gene_counts[1, ]),
  Percentage = as.numeric(gene_percentages[1, ])
) %>%
  arrange(desc(Percentage)) %>%
  select(Gene, Count, Percentage)

# 2) ARG summary table overall and by source, n (%)
tableS11 <- arg_data %>%
  pivot_longer(
    cols = all_of(arg_gene_cols),
    names_to = "Gene",
    values_to = "Present"
  ) %>%
  group_by(Gene) %>%
  summarise(
    total_n = sum(!is.na(Present)),
    clinical_n = sum(Source == "clinics" & !is.na(Present)),
    wastewater_n = sum(Source == "wastewater" & !is.na(Present)),
    .groups = "drop"
  ) %>%
  mutate(
    `Overall isolates, n (%)` = paste0(
      total_n,
      " (",
      round(100 * total_n / nrow(arg_data), 1),
      "%)"
    ),
    `Clinical isolates, n (%)` = paste0(
      clinical_n,
      " (",
      round(100 * clinical_n / sum(arg_data$Source == "clinics"), 1),
      "%)"
    ),
    `Wastewater isolates, n (%)` = paste0(
      wastewater_n,
      " (",
      round(100 * wastewater_n / sum(arg_data$Source == "wastewater"), 1),
      "%)"
    )
  ) %>%
  arrange(desc(total_n), Gene) %>%
  select(
    Gene,
    `Overall isolates, n (%)`,
    `Clinical isolates, n (%)`,
    `Wastewater isolates, n (%)`
  )

tableS11 %>%
  knitr::kable(
    format = "html",
    booktabs = TRUE,
    caption = paste0(
  "Table S11: Distribution of antimicrobial resistance genes (ARGs) among MRSA isolates recovered from clinical and wastewater sources. Values indicate the number and percentage of isolates carrying each ARG, overall and stratified by source. ARGs were identified using the Comprehensive Antibiotic Resistance Database (CARD, accessed 25 May 2025)."
)

  ) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "left"
  )

#Optional Save
# Optional export
#write.csv(tableS11, "./tables/TableS11.csv", row.names = FALSE)

# 3) Long-format ARG summary by source for plotting
# Make sure arg_class has standardized column names only once
arg_class_clean <- arg_class %>%
  rename(
    Gene = arg,
    Class = antibiotic_class
  )

gene_counts_by_source <- arg_data %>%
  select(Source, all_of(arg_gene_cols)) %>%
  group_by(Source) %>%
  summarise(
    across(
      all_of(arg_gene_cols),
      ~ sum(!is.na(.)),
      .names = "{col}"
    ),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = -Source,
    names_to = "Gene",
    values_to = "Positive_Samples"
  ) %>%
  left_join(
    arg_class_clean,
    by = "Gene"
  ) %>%
  mutate(
    Gene = recode(
      Gene,
      "PC1_beta-lactamase_(blaZ)" = "blaZ",
      "Staphylococcus_aureus_FosB" = "fosB",
      "Staphylococcus_aureus_norA" = "norA",
      "Staphylococcys_aureus_LmrS" = "lmrS"
    )
  ) %>%
  left_join(
    arg_data %>%
      dplyr::count(Source, name = "n_isolates"),
    by = "Source"
  ) %>%
  mutate(
    Proportion = Positive_Samples / n_isolates * 100,
    Proportion = ifelse(
      Proportion >= 0.1,
      round(Proportion, 1),
      round(Proportion, 2)
    ),
    Result = paste0(
      Positive_Samples,
      "; ",
      Proportion,
      "%"
    )
  ) %>%
  group_by(Gene) %>%
  filter(any(Positive_Samples > 0)) %>%
  ungroup()

# 4) Wide table with same output as before
table_arg <- gene_counts_by_source %>%
  select(Source, Gene, Result) %>%
  pivot_wider(
    names_from = Source,
    values_from = Result
  )

colnames(table_arg) <- gsub("_", " ", colnames(table_arg))
colnames(table_arg)[-1] <- paste0(
  colnames(table_arg)[-1],
  " (n; %)"
)
```

### 5.3.2 Rarefaction ARG richness by source

Subsamples the larger clinical isolate set down to the wastewater sample size (n = 26) across 10,000 iterations, so ARG richness can be compared fairly between sources despite the sample-size imbalance.

```{r arg-rarefaction, message=FALSE, warning=FALSE, eval=TRUE}
# Rarefaction of ARG richness by source
set.seed(123)

# Number of isolates in each source
n_wastewater <- sum(arg_data$Source == "wastewater")
n_clinical <- sum(arg_data$Source == "clinics")

# ARG matrix only
arg_cols <- setdiff(
  colnames(arg_data),
  c("Sample", "Source", "NUM_FOUND")
)

# Wastewater observed richness
ww_data <- arg_data %>%
  filter(Source == "wastewater")

ww_richness <- ww_data %>%
  select(all_of(arg_cols)) %>%
  summarise(across(everything(), ~ any(!is.na(.)))) %>%
  unlist() %>%
  sum()

# Clinical observed richness
clinical_data <- arg_data %>%
  filter(Source == "clinics")

clinical_richness <- clinical_data %>%
  select(all_of(arg_cols)) %>%
  summarise(across(everything(), ~ any(!is.na(.)))) %>%
  unlist() %>%
  sum()

# Rarefy clinical isolates to wastewater sample size
n_iter <- 10000

clinical_rarefied <- replicate(
  n_iter,
  {
    sampled <- clinical_data %>%
      slice_sample(n = n_wastewater)

    sampled %>%
      select(all_of(arg_cols)) %>%
      summarise(across(everything(), ~ any(!is.na(.)))) %>%
      unlist() %>%
      sum()
  }
)

clinical_rarefied_summary <- tibble(
  observed_clinical = clinical_richness,
  observed_wastewater = ww_richness,
  rarefied_median = median(clinical_rarefied),
  rarefied_mean = mean(clinical_rarefied),
  lower_95CI = quantile(clinical_rarefied, 0.025),
  upper_95CI = quantile(clinical_rarefied, 0.975)
)

clinical_rarefied_summary

# Rarefaction summary table
rarefaction_table <- tibble(
  Comparison = c(
    "Clinical isolates (observed)",
    "Wastewater isolates (observed)",
    paste0(
      "Clinical isolates rarefied to wastewater sample size (n = ",
      n_wastewater,
      ")"
    )
  ),
  `Unique ARGs` = c(
    clinical_richness,
    ww_richness,
    paste0(
      round(median(clinical_rarefied), 1),
      " (95% CI ",
      quantile(clinical_rarefied, 0.025),
      "–",
      quantile(clinical_rarefied, 0.975),
      ")"
    )
  )
)

rarefaction_table %>%
  knitr::kable(
    format = "html",
    booktabs = TRUE,
    caption = paste0(
      "Rarefaction analysis of ARG richness. Clinical isolates (n = ",
      n_clinical,
      ") were randomly subsampled to match the number of wastewater isolates (n = ",
      n_wastewater,
      ") across ",
      format(n_iter, big.mark = ","),
      " iterations. Values represent the number of unique ARGs detected."
    )
  ) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "left"
  )
```

### 5.3.3 Rarefaction Plot

Renders Figure S3: the rarefaction distribution, showing whether the observed wastewater ARG richness falls inside or outside the range expected from clinical isolates of the same sample size.

```{r fig:figureS3, fig.cap = "Figure S3: Rarefaction analysis of unique antimicrobial resistance genes (ARGs) detected among clinical MRSA isolates. To account for differences in sample size between sources, 26 distinct clinical isolates, matching the number of wastewater isolates, were randomly selected without replacement from the clinical dataset across 10,000 iterations. For each iteration, the number of unique ARGs detected among the selected isolates was recorded. The histogram shows the distribution of the number of unique ARGs across rarefaction iterations. The dashed vertical line indicates the observed number of unique ARGs detected among wastewater isolates (n = 26).", fig.height=4.5, fig.width=6.5, fig.lp=NULL, warning=FALSE, message=FALSE, eval=TRUE}

rarefaction_plot_data <- tibble(
  unique_args = clinical_rarefied
)

figureS3 <- ggplot(
  rarefaction_plot_data,
  aes(x = unique_args)
) +
  geom_histogram(
    binwidth = 1,
    fill = "grey80",
    color = "black"
  ) +
  geom_vline(
    xintercept = ww_richness,
    linetype = "dashed",
    linewidth = 1
  ) +
 annotate(
  "text",
  x = ww_richness - 0.5,
  y = Inf,
  label = paste0(
    "Observed among\nwastewater isolates\n(n = ",
    n_wastewater,
    "): ",
    ww_richness,
    " ARGs"
  ),
  hjust = 1,
  vjust = 1.5,
  size = 5
)+
labs(
  x = "Number of unique ARGs detected among 26 randomly \nselected clinical isolates",
  y = "Rarefaction iterations"
  ) +
  theme_bw() +
  theme(
    axis.text = element_text(
      color = "black",
      size = 12
    ),
    axis.title = element_text(
      color = "black",
      size = 14
    ),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

print(figureS3)

#ggsave("./figures/FigureS3.pdf",plot = figureS3,width = 11.69,height = 8.47,units = "in")
```

### 5.3.4 Isolates resistant to classes (tab2_part)

Produces the resistance-class n (%) breakdown that is part 1 of 3 of the chunks compiled into Table 2 of the main manuscript (see Section 1 Overview).

```{r tab2_part, warning=FALSE, eval=TRUE, message=FALSE}
arg_class_clean <- arg_class %>%
  rename(
    Gene = arg,
    Class = antibiotic_class
  )

arg_data_long_all_classes <- arg_data %>%
  pivot_longer(
    cols = -c(Sample, Source, NUM_FOUND),
    names_to = "Gene",
    values_to = "Present"
  ) %>%
  filter(!is.na(Present)) %>%
  left_join(arg_class_clean, by = "Gene") %>%
  separate_rows(Class, sep = "_") %>%
  filter(!is.na(Class))

isolates_per_source <- arg_data %>%
  dplyr::count(Source, name = "n_isolates") %>%
  deframe()

n_total <- nrow(arg_data)

tab2_part <- arg_data_long_all_classes %>%
  distinct(Sample, Source, Class) %>%
  group_by(Class) %>%
  summarise(
    total_n = n_distinct(Sample),
    clinical_n = n_distinct(Sample[Source == "clinics"]),
    wastewater_n = n_distinct(Sample[Source == "wastewater"]),
    .groups = "drop"
  ) %>%
  mutate(
    `Total isolates, n (%)` = paste0(
      total_n, " (", round(100 * total_n / n_total, 1), "%)"
    ),
    `Clinical isolates, n (%)` = paste0(
      clinical_n, " (", round(100 * clinical_n / isolates_per_source["clinics"], 1), "%)"
    ),
    `Wastewater isolates, n (%)` = paste0(
      wastewater_n, " (", round(100 * wastewater_n / isolates_per_source["wastewater"], 1), "%)"
    )
  ) %>%
  mutate(
    `Drug class` = recode(
      Class,
      "aminoglycoside" = "Aminoglycosides",
      "aminoglycosides" = "Aminoglycosides",
      "macrolide" = "Macrolides",
      "lincosamide" = "Lincosamides",
      "streptogramin" = "Streptogramins",
      "penam" = "Penams",
      "nucleoside" = "Nucleosides",
      "fosfomycin" = "Fosfomycin",
      "diaminopyrimidine" = "Diaminopyrimidines",
      "fusidicacid" = "Fusidic acid",
      "oxazolidinone" = "Oxazolidinones",
      "phenicol" = "Phenicols",
      "pleuromutilin" = "Pleuromutilins",
      "tetracycline" = "Tetracyclines",
      "mupirocin" = "Mupirocin",
      "effluxpump" = "Efflux pumps",
      .default = Class
    )
  ) %>%
  select(
    `Drug class`,
    `Total isolates, n (%)`,
    `Clinical isolates, n (%)`,
    `Wastewater isolates, n (%)`
  ) %>%
  arrange(desc(as.numeric(gsub(" .*", "", `Total isolates, n (%)`))))

tab2_part %>%
  knitr::kable(
    format = "html",
    booktabs = TRUE,
    caption = paste0(
      "Table 2 (part 1 of 3): Distribution of MRSA isolates by antimicrobial resistance class inferred from ARGs detected in the sequencing data, overall and by source. Feeds the resistance-class rows of Table 2 in the main manuscript; see Section 1 Overview."
    )
  ) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "left"
  )

# Optional export
#write.csv(table2_part, "./tables/tab2_part.csv", row.names = FALSE)
```

## 5.4 ARGs in wastewater and clinics (fig2)

Renders Figure 2: per-gene carriage proportions by source — the main-text visualization of which resistance genes distinguish wastewater from clinical isolates. (Renumbered to follow 5.3 sequentially.)

```{r fig:figure2, fig.cap = "Figure 2: Distribution of antibiotic resistance genes across sources in MRSA isolates. The proportion of MRSA isolates carrying genes associated with different antibiotic resistance classes, based on annotations from the Comprehensive Antibiotic Resistance Database (CARD). Source is represented in colors: clinics (pink) and wastewater (blue). *AAC(6')-APH(2'') refers to the bifunctional aminoglycoside-modifying enzyme AAC(6')-Ie-APH(2'')-Ia. Each bar shows the proportion of isolates from a specific source carrying the corresponding gene.", fig.height=8.47, fig.width=11.69, fig.lp=NULL, warning=FALSE, message=FALSE, eval=TRUE}
# Prepare ARG data
gene_counts_by_source_all <- gene_counts_by_source %>%
  filter(!is.na(Class)) %>%
  separate_rows(Class, sep = "_") %>%
  mutate(
    Class = recode(
      Class,
      "aminoglycoside" = "Aminoglycosides",
      "aminoglycosides" = "Aminoglycosides",
      "macrolide" = "Macrolides",
      "lincosamide" = "Lincosamides",
      "streptogramin" = "Streptogramins",
      "penam" = "Beta-lactams",
      "nucleoside" = "Nucleosides",
      "fosfomycin" = "Fosfomycin",
      "diaminopyrimidine" = "Diaminopyrimidines",
      "fusidicacid" = "Fusidic acid",
      "oxazolidinone" = "Oxazolidinones",
      "phenicol" = "Phenicols",
      "pleuromutilin" = "Pleuromutilins",
      "tetracycline" = "Tetracyclines",
      "mupirocin" = "Mupirocin",
      "effluxpump" = "Efflux pumps",
      .default = Class
    )
  ) %>%
  mutate(
    Class = case_when(
      Class %in% c(
        "Macrolides",
        "Lincosamides",
        "Streptogramins"
      ) ~ "Macrolides/Lincosamides/\nStreptogramins",
      TRUE ~ Class
    ),
    Gene = recode(
      Gene,
      "AAC(6')-Ie-APH(2'')-Ia" = "AAC(6')-APH(2'')*"
    )
  )

# Order genes
gene_order_all <- gene_counts_by_source_all %>%
  group_by(Gene) %>%
  summarise(
    max_prop = max(Proportion, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(max_prop)) %>%
  pull(Gene)

gene_counts_by_source_all <- gene_counts_by_source_all %>%
  mutate(
    Gene = factor(Gene, levels = gene_order_all)
  )

# Order classes
preferred_class_order <- c(
  "Beta-lactams",
  "Aminoglycosides",
  "Diaminopyrimidines",
  "Efflux pumps",
  "Fosfomycin",
  "Fusidic acid",
  "Macrolides/Lincosamides/\nStreptogramins",
  "Mupirocin",
  "Nucleosides",
  "Oxazolidinones",
  "Phenicols",
  "Pleuromutilins",
  "Tetracyclines"
)

all_class_order <- c(
  preferred_class_order,
  setdiff(
    sort(unique(gene_counts_by_source_all$Class)),
    preferred_class_order
  )
)

gene_counts_by_source_all <- gene_counts_by_source_all %>%
  mutate(
    Class = factor(Class, levels = all_class_order)
  )

# Colors
custom_colors_friendly <- c(
  "clinics" = "#CC79A7",
  "wastewater" = "#56B4E9"
)

# Plot
figure2 <- ggplot(
  gene_counts_by_source_all,
  aes(
    y = Proportion,
    x = Gene,
    fill = Source
  )
) +
  geom_bar(
    stat = "identity",
    position = position_dodge(width = 0.7),
    width = 0.6
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_text(
      hjust = 1,
      color = "black",
      size = 12
    ),
    axis.text.x = element_text(
      hjust = 1,
      color = "black",
      size = 12,
      angle = 45
    ),
    axis.title = element_text(
      size = 14,
      color = "black"
    ),
    strip.text = element_text(size = 13),
    strip.background = element_rect(
      fill = "white",
      color = "black"
    ),
    legend.position = c(0.80, 0.1),
    legend.justification = c("right", "bottom"),
    legend.direction = "vertical",   # make legend items vertical
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(
    y = "Proportion of isolates carrying the gene (%)",
    x = "",
    fill = "Source"
  ) +
  scale_fill_manual(
    values = custom_colors_friendly,
    guide = guide_legend(ncol = 1)  # ensure vertical stacking
  ) +
  facet_wrap(
    ~ Class,
    scales = "free_x",
    ncol = 5
  )

print(figure2)

#ggsave("./figures/Figure2.pdf",plot = figure2,width = 11.69,height = 8.27,units = "in")
```

## 5.5 Abundance of antimicrobial resistance determinants (figS2)

Renders Figure S2 and runs the Mann-Whitney U tests comparing ARG counts and resistance-class counts between sources, quantifying whether the visual difference in Figure 2 is statistically supported. (Renumbered to follow 5.4 sequentially.)

```{r fig:figureS2, fig.cap = "Number of antimicrobial resistance genes (ARGs) and antibiotic resistance classes detected in clinical and wastewater MRSA isolates. Resistance to an antibiotic class was defined as the presence of at least one ARG associated with resistance to that class, irrespective of gene expression. ARGs classified as efflux pumps were included in comparisons of ARG counts per isolate but excluded from comparisons of antibiotic resistance classes. Points represent individual isolates; boxes indicate the median and interquartile range. P-values were calculated using two-sided Mann-Whitney U tests.", fig.height=8.47, fig.width=11.69, fig.lp=NULL, warning=FALSE, message=FALSE, eval=TRUE}
# Prepare long-format ARG data
arg_data_long_for_counts <- arg_data %>%
  pivot_longer(
    cols = -c(Sample, Source, NUM_FOUND),
    names_to = "Gene",
    values_to = "Present"
  ) %>%
  filter(!is.na(Present)) %>%
  left_join(
    arg_class_clean,
    by = "Gene"
  ) %>%
  separate_rows(Class, sep = "_") %>%
  filter(!is.na(Class)) %>%
  mutate(
    Class = recode(
      Class,
      "aminoglycoside" = "Aminoglycosides",
      "aminoglycosides" = "Aminoglycosides",
      "macrolide" = "Macrolides",
      "lincosamide" = "Lincosamides",
      "streptogramin" = "Streptogramins",
      "penam" = "Beta-lactams",
      "nucleoside" = "Nucleosides",
      "fosfomycin" = "Fosfomycin",
      "diaminopyrimidine" = "Diaminopyrimidines",
      "fusidicacid" = "Fusidic acid",
      "oxazolidinone" = "Oxazolidinones",
      "phenicol" = "Phenicols",
      "pleuromutilin" = "Pleuromutilins",
      "tetracycline" = "Tetracyclines",
      "mupirocin" = "Mupirocin",
      "effluxpump" = "Efflux pumps",
      .default = Class
    )
  )
# Count ARGs per isolate
arg_counts <- arg_data_long_for_counts %>%
  distinct(
    Sample,
    Source,
    Gene
  ) %>%
  group_by(
    Sample,
    Source
  ) %>%
  summarise(
    ARGs = n(),
    .groups = "drop"
  )

# Count antibiotic classes per isolate
# Excludes efflux pumps
ab_class_counts <- arg_data_long_for_counts %>%
  filter(Class != "Efflux pumps") %>%
  distinct(
    Sample,
    Source,
    Class
  ) %>%
  group_by(
    Sample,
    Source
  ) %>%
  summarise(
    AB_classes = n(),
    .groups = "drop"
  )

# Merge counts
counts_summary <- arg_counts %>%
  left_join(
    ab_class_counts,
    by = c("Sample", "Source")
  )

# Reshape for plotting
counts_long <- counts_summary %>%
  pivot_longer(
    cols = c(ARGs, AB_classes),
    names_to = "Measure",
    values_to = "Count"
  ) %>%
  mutate(
    Measure = recode(
      Measure,
      ARGs = "Antimicrobial resistance genes",
      AB_classes = "Antibiotic classes"
    )
  )

# Statistical tests
stat_arg_source <- wilcox.test(
  Count ~ Source,
  data = counts_long %>%
    filter(
      Measure == "Antimicrobial resistance genes"
    )
)

stat_ab_source <- wilcox.test(
  Count ~ Source,
  data = counts_long %>%
    filter(
      Measure == "Antibiotic classes"
    )
)

# P-value labels
pval_labels <- tibble(
  Measure = c(
    "Antimicrobial resistance genes",
    "Antibiotic classes"
  ),
  label = c(
    paste0(
      "Mann-Whitney U test: p = ",
      signif(stat_arg_source$p.value, 2)
    ),
    paste0(
      "Mann-Whitney U test: p = ",
      signif(stat_ab_source$p.value, 2)
    )
  )
)

label_positions <- counts_long %>%
  group_by(Measure) %>%
  summarise(
    x = 0.65,
    y = max(Count, na.rm = TRUE) * 1.08,
    .groups = "drop"
  ) %>%
  left_join(
    pval_labels,
    by = "Measure"
  )

# Plot
figureS2 <- ggplot(
  counts_long,
  aes(
    x = Source,
    y = Count,
    fill = Source
  )
) +
  geom_boxplot(
    fill = NA,
    color = "black",
    outlier.shape = NA,
    linewidth = 0.7
  ) +
  geom_jitter(
    aes(color = Source),
    width = 0.18,
    height = 0,
    alpha = 0.75,
    size = 4
  ) +
  geom_text(
    data = label_positions,
    aes(
      x = x,
      y = y,
      label = label
    ),
    inherit.aes = FALSE,
    hjust = 0,
    vjust = 1,
    size = 6
  ) +
  scale_color_manual(
    values = custom_colors_friendly,
    name = "Source"
  ) +
  scale_fill_manual(
    values = custom_colors_friendly,
    guide = "none"
  ) +
  facet_wrap(
    ~ Measure,
    scales = "free_y"
  ) +
  scale_y_continuous(
    expand = expansion(
      mult = c(0.05, 0.20)
    )
  ) +
  labs(
    title = "",
    x = NULL,
    y = "Number of antibiotic classes/ARGs"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(
      color = "black",
      size = 20,
      angle = 45,
      hjust = 1
    ),
    axis.text.y = element_text(
      color = "black",
      size = 18
    ),
    axis.title.y = element_text(
      size = 22,
      color = "black"
    ),
    strip.text = element_text(
      size = 22,
      face = "bold",
      color = "black"
    ),
    strip.background = element_rect(
      fill = "white",
      color = "black"
    ),
    legend.position = "none",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

print(figureS2)

#ggsave("./figures/FigureS2.pdf",plot = figureS2,width = 11.69,height = 8.27,units = "in")

# Summary statistics for Figure S2, added to table 2 in main text
figureS2_summary <- counts_summary %>%
  pivot_longer(
    cols = c(ARGs, AB_classes),
    names_to = "Measure",
    values_to = "Count"
  ) %>%
  mutate(
    Measure = recode(
      Measure,
      ARGs = "Antimicrobial resistance genes",
      AB_classes = "Antibiotic classes"
    )
  ) %>%
  group_by(Measure, Source) %>%
  summarise(
    Median = median(Count),
    Q1 = quantile(Count, 0.25),
    Q3 = quantile(Count, 0.75),
    .groups = "drop"
  ) %>%
  mutate(
    `Median (IQR)` = paste0(
      Median,
      " (",
      Q1,
      "–",
      Q3,
      ")"
    )
  ) %>%
  select(
    Measure,
    Source,
    `Median (IQR)`
  ) %>%
  pivot_wider(
    names_from = Source,
    values_from = `Median (IQR)`
  )

# Overall values
overall_summary <- counts_summary %>%
  pivot_longer(
    cols = c(ARGs, AB_classes),
    names_to = "Measure",
    values_to = "Count"
  ) %>%
  mutate(
    Measure = recode(
      Measure,
      ARGs = "Antimicrobial resistance genes",
      AB_classes = "Antibiotic classes"
    )
  ) %>%
  group_by(Measure) %>%
  summarise(
    Median = median(Count),
    Q1 = quantile(Count, 0.25),
    Q3 = quantile(Count, 0.75),
    .groups = "drop"
  ) %>%
  mutate(
    Overall = paste0(
      Median,
      " (",
      Q1,
      "–",
      Q3,
      ")"
    )
  ) %>%
  select(
    Measure,
    Overall
  )

figureS2_summary <- overall_summary %>%
  left_join(
    figureS2_summary,
    by = "Measure"
  )

figureS2_summary %>%
  knitr::kable(
    format = "html",
    booktabs = TRUE,
    caption = "Table 2 (part 2 of 3): Median number of antimicrobial resistance genes (ARGs) and antibiotic resistance classes detected per MRSA isolate overall and by source. Values are reported as median (IQR). Feeds the ARG/class median rows of Table 2 in the main manuscript; see Section 1 Overview."
  ) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "left"
  )
```

## 5.6 Multidrug resistance (tabS10)

Produces Table S10: the distribution of isolates by number of resistance classes carried, and the overall MDR (≥3 classes) rate — the summary statistic most often quoted in the Results/Discussion. (Renumbered to follow 5.5 sequentially.)

```{r tableS10, message=FALSE, warning=FALSE, eval=TRUE}
# MDR summary by number of antibiotic classes (Efflux pump genes/classes excluded)
# Standardize ARG class annotation
arg_class_clean <- arg_class %>%
  rename(
    Gene = arg,
    Class = antibiotic_class
  )

# Define classes included in MDR definition
included_classes <- c(
  "aminoglycoside",
  "macrolide",
  "lincosamide",
  "streptogramin",
  "penam",
  "nucleoside",
  "fosfomycin",
  "diaminopyrimidine",
  "fusidicacid",
  "oxazolidinone",
  "phenicol",
  "pleuromutilin",
  "tetracycline",
  "mupirocin"
)

# Identify ARGs that are NOT exclusively efflux pumps
non_efflux_genes <- arg_class_clean %>%
  separate_rows(Class, sep = "_") %>%
  group_by(Gene) %>%
  summarise(
    has_non_efflux_class = any(Class %in% included_classes),
    .groups = "drop"
  ) %>%
  filter(has_non_efflux_class) %>%
  pull(Gene)

# Long-format ARG data with efflux-only genes excluded
arg_data_long_mdr <- arg_data %>%
  pivot_longer(
    cols = -c(Sample, Source, NUM_FOUND),
    names_to = "Gene",
    values_to = "Present"
  ) %>%
  filter(
    !is.na(Present),
    Gene %in% non_efflux_genes
  ) %>%
  left_join(
    arg_class_clean,
    by = "Gene"
  ) %>%
  separate_rows(Class, sep = "_") %>%
  filter(Class %in% included_classes)

# Count number of unique included antibiotic classes per isolate
mdr_counts <- arg_data_long_mdr %>%
  distinct(
    Sample,
    Source,
    Class
  ) %>%
  group_by(
    Sample,
    Source
  ) %>%
  summarise(
    `Number of drug classes` = n(),
    .groups = "drop"
  )

# Add isolates with zero detected included classes, if any
all_samples <- arg_data %>%
  distinct(
    Sample,
    Source
  )

mdr_counts <- all_samples %>%
  left_join(
    mdr_counts,
    by = c("Sample", "Source")
  ) %>%
  mutate(
    `Number of drug classes` = ifelse(
      is.na(`Number of drug classes`),
      0,
      `Number of drug classes`
    )
  )

# Denominators
n_total <- n_distinct(mdr_counts$Sample)
n_clinics <- sum(mdr_counts$Source == "clinics")
n_wastewater <- sum(mdr_counts$Source == "wastewater")

# Helper function for n (%) formatting
format_n_pct <- function(n, denom) {
  paste0(
    n,
    " (",
    round(100 * n / denom, 1),
    "%)"
  )
}

# Summary by number of drug classes
mdr_summary <- mdr_counts %>%
  group_by(`Number of drug classes`) %>%
  summarise(
    overall_n = n(),
    clinics_n = sum(Source == "clinics"),
    wastewater_n = sum(Source == "wastewater"),
    .groups = "drop"
  ) %>%
  mutate(
    `Number of drug classes` = as.character(`Number of drug classes`),
    `Overall isolates, n (%)` = format_n_pct(overall_n, n_total),
    `Clinical isolates, n (%)` = format_n_pct(clinics_n, n_clinics),
    `Wastewater isolates, n (%)` = format_n_pct(wastewater_n, n_wastewater)
  ) %>%
  select(
    `Number of drug classes`,
    `Overall isolates, n (%)`,
    `Clinical isolates, n (%)`,
    `Wastewater isolates, n (%)`
  )

# MDR = 3+ row
mdr_3plus_counts <- mdr_counts %>%
  summarise(
    overall_n = sum(`Number of drug classes` >= 3),
    clinics_n = sum(Source == "clinics" & `Number of drug classes` >= 3),
    wastewater_n = sum(Source == "wastewater" & `Number of drug classes` >= 3)
  ) %>%
  mutate(
    `Number of drug classes` = "MDR ≥3",
    `Overall isolates, n (%)` = format_n_pct(overall_n, n_total),
    `Clinical isolates, n (%)` = format_n_pct(clinics_n, n_clinics),
    `Wastewater isolates, n (%)` = format_n_pct(wastewater_n, n_wastewater)
  ) %>%
  select(
    `Number of drug classes`,
    `Overall isolates, n (%)`,
    `Clinical isolates, n (%)`,
    `Wastewater isolates, n (%)`
  )

# Final MDR table
tableS10 <- bind_rows(
  mdr_summary,
  mdr_3plus_counts
)

# Display table
tableS10 %>%
  knitr::kable(
    format = "html",
    booktabs = TRUE,
    caption = paste0(
      "Table S10: Distribution of MRSA isolates by number of antimicrobial resistance classes inferred from ARGs detected in the sequencing data. Values indicate the number and percentage of isolates carrying ARGs associated with each number of resistance classes, based on Comprehensive Antibiotic Resistance Database (CARD) annotations. Multidrug resistance (MDR) was defined as the presence of ARGs associated with resistance to three or more antimicrobial classes. The “efflux pumps” category was not considered an antimicrobial resistance class."
    )
  ) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "left"
  )

# Optional export
#write.csv(tableS10, "./tables/TableS10.csv", row.names = FALSE)
```

------------------------------------------------------------------------

# 6. Virulence Factor Genes (VFs)

Characterizes virulence factor gene carriage, most importantly Panton-Valentine leukocidin (PVL) status, which also determines the red-dot markers added to Figure 1. Feeds Table S12 and part of Table 2.

## 6.1 Load and format data

Loads the raw Abricate virulence-factor hits per isolate and assigns each gene to a predefined VF category (PVL, enterotoxins, exfoliative toxins, other) used throughout this section.

```{r vfs-clean-analysis, message=FALSE, warning=FALSE, eval=TRUE}
# Load and format VF data
vf_data <- read.delim(
  "data/vfs_abricate.txt",
  header = TRUE,
  check.names = FALSE
)

colnames(vf_data)[1] <- "Sample"

vf_data$Sample <- gsub("^.*/|_.*$", "", vf_data$Sample)
vf_data[vf_data == "."] <- NA

vf_data <- vf_data %>%
  rename(Sample_ID = Sample) %>%
  left_join(
    sample_metadata %>%
      select(Sample_ID, Source, Sample_date),
    by = "Sample_ID"
  ) %>%
  rename(Sample = Sample_date) %>%
  relocate(Sample, Source) %>%
  select(-Sample_ID) %>%
  filter(!is.na(Sample), !is.na(Source))

# Define VF categories
known_vf_categories <- tibble(
  Gene = c(
    "lukF-PV",
    "lukS-PV",
    "sea",
    "seb",
    "sec",
    "sed",
    "seh",
    "selk",
    "sell",
    "selq",
    "hly/hla",
    "hlb",
    "hld",
    "hlgA",
    "hlgB",
    "hlgC",
    "eta",
    "etb",
    "spa"
  ),
  `VF category` = c(
    rep("Panton-Valentine leukocidin", 2),
    rep("Staphylococcal enterotoxins", 8),
    rep("Hemolysins", 6),
    rep("Exfoliative toxins", 2),
    "Staphylococcal protein A"
  )
)

# Identify all VF gene columns
vf_gene_cols <- setdiff(
  colnames(vf_data),
  c("Sample", "Source", "NUM_FOUND")
)

# Assign predefined categories; all other genes become "Other virulence factors"
vf_categories <- tibble(
  Gene = vf_gene_cols
) %>%
  left_join(
    known_vf_categories,
    by = "Gene"
  ) %>%
  mutate(
    `VF category` = ifelse(
      is.na(`VF category`),
      "Other virulence factors",
      `VF category`
    )
  )
```

## 6.2 All genes (tabS12)

Produces Table S12: carriage of every individual virulence factor gene, including the combined lukF-PV + lukS-PV PVL-positive row (≥80% identity for both genes) — the same ≥80%-identity, both-genes rule used to decide which isolates receive the red PVL marker in Figure 1.

```{r tableS12, eval=TRUE, message=FALSE}
# Long-format VF data
vf_long <- vf_data %>%
  select(Sample, Source, all_of(vf_gene_cols)) %>%
  pivot_longer(
    cols = all_of(vf_gene_cols),
    names_to = "Gene",
    values_to = "Value"
  ) %>%
  mutate(
    Identity = suppressWarnings(
      as.numeric(str_split(Value, ";", simplify = TRUE)[, 1])
    ),
    Present = !is.na(Identity) & Identity >= 80
  ) %>%
  left_join(
    vf_categories,
    by = "Gene"
  )

# Denominators
isolates_per_source <- vf_data %>%
  distinct(Sample, Source) %>%
  dplyr::count(Source, name = "n") %>%
  deframe()

n_clinical <- as.numeric(isolates_per_source["clinics"])
n_wastewater <- as.numeric(isolates_per_source["wastewater"])
n_total <- n_clinical + n_wastewater

format_n_pct <- function(n, denom) {
  paste0(
    n,
    "/",
    denom,
    " (",
    round(100 * n / denom, 1),
    "%)"
  )
}

# Supplementary VF gene table
category_order <- c(
  "Panton-Valentine leukocidin",
  "Staphylococcal enterotoxins",
  "Hemolysins",
  "Exfoliative toxins",
  "Staphylococcal protein A",
  "Other virulence factors"
)

vf_gene_summary <- vf_long %>%
  group_by(`VF category`, Gene) %>%
  summarise(
    overall_n = sum(Present, na.rm = TRUE),
    clinical_n = sum(Source == "clinics" & Present, na.rm = TRUE),
    wastewater_n = sum(Source == "wastewater" & Present, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    `VF category` = factor(`VF category`, levels = category_order),
    `Overall isolates, n/N (%)` = format_n_pct(overall_n, n_total),
    `Clinical isolates, n/N (%)` = format_n_pct(clinical_n, n_clinical),
    `Wastewater isolates, n/N (%)` = format_n_pct(wastewater_n, n_wastewater)
  ) %>%
  arrange(`VF category`, desc(overall_n), Gene) %>%
  select(
    `VF category`,
    Gene,
    `Overall isolates, n/N (%)`,
    `Clinical isolates, n/N (%)`,
    `Wastewater isolates, n/N (%)`
  )

# Add PVL-positive row: isolates carrying both lukF-PV and lukS-PV
pvl_both_summary <- vf_long %>%
  filter(Gene %in% c("lukF-PV", "lukS-PV")) %>%
  group_by(Sample, Source) %>%
  summarise(
    PVL_positive = all(Present),
    .groups = "drop"
  ) %>%
  summarise(
    overall_n = sum(PVL_positive, na.rm = TRUE),
    clinical_n = sum(Source == "clinics" & PVL_positive, na.rm = TRUE),
    wastewater_n = sum(Source == "wastewater" & PVL_positive, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    `VF category` = factor(
      "Panton-Valentine leukocidin",
      levels = category_order
    ),
    Gene = "lukF-PV + lukS-PV",
    `Overall isolates, n/N (%)` = format_n_pct(overall_n, n_total),
    `Clinical isolates, n/N (%)` = format_n_pct(clinical_n, n_clinical),
    `Wastewater isolates, n/N (%)` = format_n_pct(wastewater_n, n_wastewater)
  ) %>%
  select(
    `VF category`,
    Gene,
    `Overall isolates, n/N (%)`,
    `Clinical isolates, n/N (%)`,
    `Wastewater isolates, n/N (%)`
  )

tableS12 <- bind_rows(
  pvl_both_summary,
  vf_gene_summary
) %>%
  arrange(`VF category`, desc(Gene == "lukF-PV + lukS-PV"), Gene)

# Display table
tableS12 %>%
  knitr::kable(
    format = "html",
    booktabs = TRUE,
    caption = paste0(
      "Table S12:  Distribution of virulence factor genes among MRSA isolates. Values indicate the number and percentage of isolates carrying each gene and are reported as n/N (%), where N is the number of isolates within the corresponding category. Genes were considered present when sequence identity was ≥80%. The row lukF-PV + lukS-PV indicates isolates carrying both PVL subunit genes. Virulence factor genes not assigned to the predefined categories were grouped as other virulence factors."
    )
  ) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "left"
  )

# Optional export
#write.csv(tableS12, "./tables/TableS12.csv", row.names = FALSE)
```

## 6.3 Major categories

Collapses the gene-level results into isolate-level positivity per VF category (PVL, enterotoxins, exfoliative toxins, spa) — part 3 of 3 of the chunks compiled into Table 2 of the main manuscript (see Section 1 Overview). (Renumbered from the original duplicate "6.2" heading.)

```{r virulence-summary-table, message=FALSE, warning=FALSE, eval=TRUE}
# Virulence summary table for main manuscript (part of Table 2)
# PVL (both genes present)
pvl_summary <- vf_data %>%
  mutate(
    lukF = as.numeric(str_split(`lukF-PV`, ";", simplify = TRUE)[,1]),
    lukS = as.numeric(str_split(`lukS-PV`, ";", simplify = TRUE)[,1]),
    Positive = !is.na(lukF) &
      !is.na(lukS) &
      lukF >= 80 &
      lukS >= 80
  )

# Enterotoxins
enterotoxin_genes <- c(
  "sea", "seb", "sec", "sed",
  "seh", "selk", "sell", "selq"
)

enterotoxin_summary <- vf_data %>%
  mutate(
    Positive = apply(
      select(., all_of(enterotoxin_genes)),
      1,
      function(x) {
        any(
          suppressWarnings(
            as.numeric(
              str_split(x, ";", simplify = TRUE)[,1]
            )
          ) >= 80,
          na.rm = TRUE
        )
      }
    )
  )

# Exfoliative toxins
exfoliative_summary <- vf_data %>%
  mutate(
    eta_present = suppressWarnings(
      as.numeric(str_split(eta, ";", simplify = TRUE)[,1])
    ) >= 80,
    etb_present = suppressWarnings(
      as.numeric(str_split(etb, ";", simplify = TRUE)[,1])
    ) >= 80,
    Positive = eta_present | etb_present
  )

# spa
spa_summary <- vf_data %>%
  mutate(
    Positive =
      suppressWarnings(
        as.numeric(str_split(spa, ";", simplify = TRUE)[,1])
      ) >= 80
  )

# Helper function
summarise_trait <- function(data, trait_name) {

  overall_n <- sum(data$Positive, na.rm = TRUE)

  clinical_n <- sum(
    data$Source == "clinics" &
      data$Positive,
    na.rm = TRUE
  )

  wastewater_n <- sum(
    data$Source == "wastewater" &
      data$Positive,
    na.rm = TRUE
  )

  tibble(
    Characteristic = trait_name,
    Overall = paste0(
      overall_n, "/", n_total,
      " (",
      round(100 * overall_n / n_total, 1),
      "%)"
    ),
    Clinical = paste0(
      clinical_n, "/", n_clinical,
      " (",
      round(100 * clinical_n / n_clinical, 1),
      "%)"
    ),
    Wastewater = paste0(
      wastewater_n, "/", n_wastewater,
      " (",
      round(100 * wastewater_n / n_wastewater, 1),
      "%)"
    )
  )
}

virulence_summary_table <- bind_rows(
  summarise_trait(
    pvl_summary,
    "PVL positive (lukF-PV + lukS-PV)"
  ),
  summarise_trait(
    enterotoxin_summary,
    "≥1 enterotoxin gene"
  ),
  summarise_trait(
    exfoliative_summary,
    "Exfoliative toxin positive (eta/etb)"
  ),
  summarise_trait(
    spa_summary,
    "spa positive"
  )
)

virulence_summary_table %>%
  knitr::kable(
    format = "html",
    booktabs = TRUE,
    caption = paste0(
      "Table 2 (part 3 of 3): Distribution of major virulence determinants (PVL, enterotoxins, exfoliative toxins, spa) among MRSA isolates. ",
      "Values are reported as n/N (%), where N is the number of isolates within ",
      "the corresponding category. Genes were considered present when sequence ",
      "identity was ≥80%. Feeds the virulence-factor rows of Table 2 in the main manuscript; see Section 1 Overview."
    )
  ) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "left"
  )
```

# 7. Genetic similarity SNPs

Uses the core-genome SNP distance matrix to identify near-identical and closely related isolate pairs, both to remove technical replicates of the same sample and to flag likely transmission-linked pairs across source compartments. Feeds Table S2, Table 1, and TableSX.

## 7.1 Load SNPs data

Loads the pairwise, recombination-masked core-genome SNP distance matrix underlying every comparison in this section.

```{r data-snp, warning=FALSE, message=FALSE, eval=TRUE}
snps_data <- read.table("data/distances_no_recomb.tab", header = TRUE, sep = "\t", check.names = FALSE)
```

## 7.2 Format SNP matrix using metadata

Relabels the distance matrix's technical sample IDs with the human-readable Sample_date identifiers used elsewhere, so downstream pair lists are directly interpretable.

```{r rename-data-snp, warning=FALSE, message=FALSE, eval=TRUE}
# Create a lookup vector from Sample_ID to Sample
name_lookup <- setNames(sample_metadata$Sample_date, sample_metadata$Sample_ID)

# Extract the matrix from snps_data and set row/column names
distance_matrix <- as.matrix(snps_data[,-1])  # remove first column
rownames(distance_matrix) <- snps_data[[1]]

colnames(distance_matrix) <- snps_data[[1]]

clean_names <- sub("_[0-9]+$", "", rownames(distance_matrix))
rownames(distance_matrix) <- clean_names
colnames(distance_matrix) <- clean_names

# Rename rows and columns using the lookup
new_names <- name_lookup[rownames(distance_matrix)]

# Check for any unmatched names
if (any(is.na(new_names))) {
  warning("Some Sample_IDs from SNP matrix were not found in sample_metadata!")
}

# Assign new names
rownames(distance_matrix) <- new_names
colnames(distance_matrix) <- new_names
```

## 7.3 Collapse identical sequences from the same sample (tabS2)

Identifies isolates with 0 cgSNPs recovered from the same sample and collapses each group to a single representative, producing Table S2 — this avoids double-counting technical replicates as independent isolates in every other analysis in this notebook.

Refined logic for defining isolates collected from the "same sample" (per source):

| Source     | **Rule for grouping isolates**                        |
|------------|-------------------------------------------------------|
| wastewater | Use `Location + Date + Source_detailed`               |
| clinics    | Each isolate is from a different sample → no grouping |

```{r tableS2, warning=FALSE, message=FALSE, eval=TRUE}
# Define Sample_Group using refined rules
sample_metadata <- sample_metadata %>%
  mutate(Sample_Group = case_when(
    Source == "wastewater" ~ paste(Location, Date, Source_detailed, sep = "_"),
    TRUE ~ Sample  # for clinical and non-replicate animal isolates
  ))

# Create lookup for sample groups
sample_groups <- sample_metadata$Sample_Group
names(sample_groups) <- sample_metadata$Sample

# Convert SNP matrix to long format and filter identical (0 SNP) pairs
distance_long <- melt(as.matrix(distance_matrix), varnames = c("Isolate1", "Isolate2"), value.name = "SNPs") %>%
  filter(Isolate1 != Isolate2 & SNPs == 0)

# Add group info and keep pairs from same group
identical_pairs <- distance_long %>%
  mutate(Group1 = sample_groups[Isolate1],
         Group2 = sample_groups[Isolate2]) %>%
  filter(Group1 == Group2)

# Create undirected graph of identical isolate pairs
edges <- identical_pairs[, c("Isolate1", "Isolate2")]
g <- igraph::graph_from_data_frame(edges, directed = FALSE)

# Extract connected components
clusters <- igraph::components(g)

# Choose one representative per group
representative_map <- data.frame(Isolate = names(clusters$membership),
                                 Cluster = clusters$membership) %>%
  group_by(Cluster) %>%
  slice(1) %>%
  ungroup()

# Identify isolates to remove
all_identical_isolates <- names(clusters$membership)
representative_isolates <- representative_map$Isolate
isolates_to_remove <- setdiff(all_identical_isolates, representative_isolates)

#  Optional: collapse the distance matrix
collapsed_isolates <- setdiff(rownames(distance_matrix), isolates_to_remove)
collapsed_isolates <- collapsed_isolates[!is.na(collapsed_isolates)]

distance_matrix_collapsed <- distance_matrix[collapsed_isolates, collapsed_isolates]

# Create full cluster membership mapping
all_membership <- data.frame(
  Isolate = names(clusters$membership),
  Cluster = clusters$membership,
  stringsAsFactors = FALSE
)

# Map each removed isolate to its retained representative
cluster_map <- left_join(all_membership, representative_map, by = "Cluster", suffix = c("_Original", "_Retained"))

# Filter out retained isolates
tableS2 <- cluster_map %>%
  filter(Isolate_Original != Isolate_Retained) %>%
  rename(Removed_Isolate = Isolate_Original, Retained_Isolate = Isolate_Retained)

# Add metadata (Source and Date)
tableS2 <- tableS2 %>%
  left_join(sample_metadata, by = c("Removed_Isolate" = "Sample")) %>%
  select(Removed_Isolate, Retained_Isolate, Source, Date) %>%
  arrange(Removed_Isolate)

# Clean column names
colnames(tableS2) <- c("Removed Isolate", "Retained Isolate", "Compartment", "Collection Date")

# Display nicely formatted table
tableS2 %>%
  kable(
    format = "html",
    longtable = TRUE,
    booktabs = TRUE,
    caption = "Table S2: Identical MRSA isolates (0 cgSNPs) originating from the same sample and collapsed into a single representative. For each case, the retained isolate is listed alongside the collapsed identical sequences, source compartment, and collection date. This includes multiple clonal isolates from the same wastewater sample."
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "repeat_header"),
    full_width = FALSE,
    position = "center"
  )

# Save to file
#write.csv(tableS2, "./tables/TableS2.csv", row.names = FALSE)
```

## 7.4 Count clonal pairs (Table1_partial)

Counts isolate pairs at or below the \<=15-cgSNP relatedness threshold, within and between source/comparison categories — the cgSNP relatedness figures that form the last chunk compiled into Table 1 of the main manuscript (see Section 1 Overview).

```{r table1_part3, warning=FALSE, message=FALSE, eval=TRUE}
# Table 1 (partial): Genetically related isolate pairs ≤15 cgSNPs
genetically_related_pairs <- distance_matrix_collapsed <= 15

compartment_vector <- sample_metadata$Source
names(compartment_vector) <- sample_metadata$Sample_date
compartment_vector <- compartment_vector[rownames(distance_matrix_collapsed)]

extract_pairs <- function(pair_matrix, compartments) {
  as.data.frame(as.table(pair_matrix)) %>%
    dplyr::filter(
      !is.na(Freq),
      Var1 != Var2,
      Freq
    ) %>%
    dplyr::mutate(
      Sample1 = pmin(as.character(Var1), as.character(Var2)),
      Sample2 = pmax(as.character(Var1), as.character(Var2)),
      Compartment1 = compartments[Sample1],
      Compartment2 = compartments[Sample2],
      Comparison = dplyr::case_when(
        Compartment1 == Compartment2 ~ Compartment1,
        Compartment1 != Compartment2 ~ paste(
          pmin(Compartment1, Compartment2),
          pmax(Compartment1, Compartment2),
          sep = " vs "
        ),
        TRUE ~ NA_character_
      )
    ) %>%
    dplyr::distinct(Sample1, Sample2, .keep_all = TRUE)
}

df_leq15 <- extract_pairs(
  genetically_related_pairs,
  compartment_vector
)

within_totals <- table(compartment_vector) %>%
  as.data.frame() %>%
  dplyr::rename(
    Comparison = compartment_vector,
    n_isolates = Freq
  ) %>%
  dplyr::mutate(
    Possible_Pairs = n_isolates * (n_isolates - 1) / 2
  ) %>%
  dplyr::select(Comparison, Possible_Pairs)

tab_compartments <- table(compartment_vector)
compartment_pairs <- t(combn(names(tab_compartments), 2))

between_totals <- data.frame(
  Comparison = paste(
    pmin(compartment_pairs[, 1], compartment_pairs[, 2]),
    pmax(compartment_pairs[, 1], compartment_pairs[, 2]),
    sep = " vs "
  ),
  Possible_Pairs =
    as.numeric(tab_compartments[compartment_pairs[, 1]]) *
    as.numeric(tab_compartments[compartment_pairs[, 2]])
)

possible_pairs <- dplyr::bind_rows(
  within_totals,
  between_totals
)

observed_pairs <- df_leq15 %>%
  dplyr::group_by(Comparison) %>%
  dplyr::summarise(
    Pairs = dplyr::n(),
    .groups = "drop"
  )

table1_partial <- observed_pairs %>%
  dplyr::left_join(
    possible_pairs,
    by = "Comparison"
  ) %>%
  dplyr::mutate(
    Percentage = round(100 * Pairs / Possible_Pairs, 2),
    `Genetically related pairs, n/total (%)` = paste0(
      Pairs,
      "/",
      Possible_Pairs,
      " (",
      format(Percentage, nsmall = 2),
      "%)"
    )
  ) %>%
  dplyr::arrange(Comparison) %>%
  dplyr::select(
    Comparison,
    `Genetically related pairs, n/total (%)`
  )

# Add total row
table1_partial <- dplyr::bind_rows(
  table1_partial,
  tibble::tibble(
    Comparison = "Total",
    `Genetically related pairs, n/total (%)` = paste0(
      nrow(df_leq15),
      "/",
      sum(possible_pairs$Possible_Pairs),
      " (",
      format(
        round(
          100 * nrow(df_leq15) /
            sum(possible_pairs$Possible_Pairs),
          2
        ),
        nsmall = 2
      ),
      "%)"
    )
  )
)

# Display Table 1 (partial)
table1_partial %>%
  knitr::kable(
    format = "html",
    booktabs = TRUE,
    escape = FALSE,
    align = c("l", "c"),
    caption = "Table 1 (partial): Genetically related MRSA isolate pairs defined by ≤15 core-genome SNPs (cgSNPs). Values are reported as the number of genetically related pairs divided by the total number of possible isolate pairs within each comparison category, with percentages shown in parentheses. Duplicate colonies with 0 cgSNPs recovered from the same wastewater sample were collapsed before analysis."
  ) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "center",
    bootstrap_options = c("striped", "condensed")
  ) %>%
  kableExtra::row_spec(
    which(tableS8$Comparison == "Total"),
    bold = TRUE,
    background = "#f0f0f0"
  )
```

## 7.5 Identify ≤15-SNP isolate pairs between source

Lists the individual isolate pairs (not just counts) that fall at or below the 15-cgSNP threshold and originate from different source compartments, exported as TableSX — an exploratory pair-level listing kept separate from, and not identical to, the summary counts in Table1_partial.

Chunk used to identify which isolates coming from different sources, namely wastewater and clinics, have 15 or less SNPs difference. Those that are exactly the same (SNP=0), are those mentioned in the manuscript.

```{r identify-between-source-15snps-pairs, warning=FALSE, message=FALSE, eval=TRUE}
# 1. Extract all isolate pairs with ≤15 SNPs (unique, off-diagonal)
snp15_pairs <- melt(
  as.matrix(distance_matrix_collapsed),
  varnames = c("Isolate1", "Isolate2"),
  value.name = "SNPs"
) %>%
  mutate(
    Isolate1 = as.character(Isolate1),
    Isolate2 = as.character(Isolate2)
  ) %>%
  filter(Isolate1 < Isolate2 & SNPs <= 15)

# 2. Add metadata for both isolates and keep only between-source pairs
snp15_metadata <- snp15_pairs %>%
  left_join(sample_metadata, by = c("Isolate1" = "Sample_date")) %>%
  rename(
    `Compartment sample 1` = Source,
    `Date sample 1` = Date,
    `Region sample 1` = Region
  ) %>%
  left_join(sample_metadata, by = c("Isolate2" = "Sample_date")) %>%
  rename(
    `Compartment sample 2` = Source,
    `Date sample 2` = Date,
    `Region sample 2` = Region
  ) %>%
  filter(`Compartment sample 1` != `Compartment sample 2`) %>%
  select(
    `Isolate 1` = Isolate1,
    `Isolate 2` = Isolate2,
    SNPs,
    `Compartment sample 1`,
    `Compartment sample 2`,
    `Date sample 1`,
    `Date sample 2`,
    `Region sample 1`,
    `Region sample 2`
  ) %>%
  mutate(
    `Date sample 1` = gsub("_", "/", `Date sample 1`),
    `Date sample 2` = gsub("_", "/", `Date sample 2`),
    `Date sample 1` = sub("\\s.*", "", `Date sample 1`),
    `Date sample 2` = sub("\\s.*", "", `Date sample 2`)
  ) %>%
  arrange(SNPs)

# Save as CSV
#write.csv(snp15_metadata,"./tables/TableSX.csv",row.names = FALSE)

# Render table
snp15_metadata %>%
  kable(
    format = "html",
    booktabs = TRUE,
    longtable = TRUE,
    escape = FALSE,
    caption = "Table SX: MRSA isolate pairs from different source compartments differing by ≤15 core-genome SNPs. For each pair, the SNP distance, source compartment, sampling date, and region are shown."
  ) %>%
  kable_styling(
    full_width = FALSE,
    position = "center"
  )
```

# 8. Map (figS1)

Renders Figure S1: the geographic map of WWTP catchments and contributing hospitals, using external shapefiles for Swiss cantons and treatment-plant catchments.

```{r fig:figureS1, fig.cap = "Figure S1: Geographic distribution of the six investigated wastewater treatment plant (WWTP) catchment areas in Switzerland. Polygons show each WWTP catchment, colored by plant, with legend entries listing the WWTP name and the estimated population served. Hospital symbols and red circles mark hospitals that provided clinical MRSA isolates, and corresponding placement within the catchment area, including Geneva University Hospitals (HUG, Geneva catchment area), Ente Ospedaliero Cantonale (EOC, samples from hospitals within the Lugano catchment), and University Hospital Zurich (USZ, Zurich catchment area). The figure was generated in R v2025.09.0 and further modified using Inkscape v1.4.2.", fig.height=8.47, fig.width=11.69, fig.lp=NULL, warning=FALSE, message=FALSE, eval=TRUE}
# ---- Global plotting style ----
theme_set(theme_void(base_size = 12))
options(sf_use_s2 = FALSE)  # avoid occasional geometry issues with CH projections

# ---- Paths (edit to your local files) ----
catchment_shp <- "data/230214_ARA_BAG.shp"
canton_shp    <- "data/swissBOUNDARIES3D_1_4_TLM_KANTONSGEBIET.shp"

# ---- Helper: safely enforce LV95 (EPSG:2056) ----
enforce_lv95 <- function(x) {
  x <- sf::st_make_valid(x)
  crs <- sf::st_crs(x)
  if (is.na(crs)) { sf::st_crs(x) <- 2056; return(x) }
  if (is.na(crs$epsg) || crs$epsg != 2056) x <- sf::st_transform(x, 2056)
  x
}

# ---- Cantons & outline ----
cantons     <- enforce_lv95(sf::read_sf(canton_shp))
switzerland <- sf::st_union(cantons)


# ---- Catchments (the six WWTPs) ----
catchments <- enforce_lv95(sf::read_sf(catchment_shp)) |>
  dplyr::select(ara_id, geometry) |>
  dplyr::mutate(ara_id = as.character(ara_id)) |>
  dplyr::filter(ara_id %in% c("323700","390101","66700","26101","515100","664301"))

# ---- Names + dummy population served (EDIT values) ----
wwtp_info <- tibble::tibble(
  ara_id = c("323700","390101","66700","26101","515100","664301"),
  wwtp_name = c("ARA Altenrhein", "ARA Chur", "ARA Sensetal-Laupen",
                "ARA Werdhölzli Zürich", "CDA Lugano", "STEP d'Aïre Genève"),
  pop_served = c(471000, 55000, 62000, 454000, 124000, 64000)
) |>
  dplyr::mutate(label = paste0(wwtp_name, " (", scales::comma(pop_served), ")"))

catchments_lab <- dplyr::left_join(catchments, wwtp_info, by = "ara_id")

# ---- Color-blind–friendly palette (Okabe–Ito) ----
okabe_ito <- c("#E69F00","#56B4E9","#009E73","#F0E442","#0072B2","#D55E00","#CC79A7","#999999")
pal <- setNames(okabe_ito[seq_len(nrow(catchments_lab))], unique(catchments_lab$label))

# ---- Map extent (pad around Switzerland) ----
bbox  <- sf::st_bbox(switzerland)
pad_x <- (bbox["xmax"] - bbox["xmin"]) * 0.03
pad_y <- (bbox["ymax"] - bbox["ymin"]) * 0.03
lims  <- list(x = c(bbox["xmin"] - pad_x, bbox["xmax"] + pad_x),
              y = c(bbox["ymin"] - pad_y, bbox["ymax"] + pad_y))

border_col <- "grey30"

# --- Controls for alignment ---
scale_width_hint <- 0.24                             # fraction of panel width
scale_pad_x      <- grid::unit(5, "cm")            # small left margin
scale_pad_y      <- grid::unit(0.1, "cm")            # detach vertically from bottom
arrow_gap        <- grid::unit(0.6, "cm")            # gap between scale and arrow

figureS1 <- ggplot() +
  # Canton polygons with dark-grey borders
  geom_sf(data = cantons, fill = "white", color = border_col, linewidth = 0.4) +
  # Country outline (same dark-grey)
  geom_sf(data = switzerland, fill = NA, color = border_col, linewidth = 0.4)

# Catchments (on top)
figureS1 <- figureS1 +
  geom_sf(
    data = catchments_lab,
    aes(fill = factor(label)),
    color = "grey30",
    linewidth = 0.4,
    alpha = 0.95
  ) +
  # --- Scale bar: left, higher off the bottom, solid black/white ---
  ggspatial::annotation_scale(
    location    = "bl",
    width_hint  = scale_width_hint,
    style       = "bar",
    bar_cols    = c("black", "white"),
    text_cex    = 0.8,
    height      = grid::unit(0.2, "cm"),
    pad_x       = scale_pad_x,
    pad_y       = scale_pad_y,
    units       = "km"
  ) +
  # --- North arrow: same vertical pad as scale, shifted right by the scale width + gap ---
  ggspatial::annotation_north_arrow(
    location    = "bl",
    which_north = "true",
    style       = ggspatial::north_arrow_fancy_orienteering,
    height      = grid::unit(1.0, "cm"),
    width       = grid::unit(1.0, "cm"),
    pad_x       = scale_pad_x + grid::unit(scale_width_hint, "npc") + arrow_gap,
    pad_y       = scale_pad_y
  ) +
  scale_fill_manual(name = "WWTP (population served)", values = pal) +
  coord_sf(xlim = lims$x, ylim = lims$y, expand = FALSE) +
  theme_void(base_size = 12) +
  theme(
    legend.position = "right",
    legend.title    = element_text(face = "bold"),
    legend.text     = element_text(size = 10),
    plot.margin     = margin(10, 12, 10, 12)
  )

figureS1
#Save
#ggsave(".figures/FigureS1.pdf", figureS1, width = 9.5, height = 6.5, units = "in")   
```
